Missing the Point\(\colon\) Non-Convergence in Multiple Imputation Procedures

### 2020-04-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Read entire research report and put comments in it for thesis manuscript. Ran 5,25,50,75,and 95% miss with 100 sims each.

At some point, any scientist conducting statistical analyses will run
into a missing data problem \citep{alli02}. Missingness is problematic
because statistical inference cannot be performed on incomplete data
without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
which may yield wildly invalid results \citep{buur18}. A popular answer
to the ubiquitous problem of missing information is to use the framework
of multiple imputation (MI), proposed by \citet{rubin87}. For the
theoretical foundation of MI, see \citet{rubin87}. For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~\citet{buur18}. With MI, missing datapoints are `imputed' (i.e.,
filled in) via an \emph{imputation model}. The imputation model can be a
joint model for all variables, or several univariate conditional
distributions (fully conditional specification). FCS is more flexible
and ``avoids the problem of specifying an appropriate joint imputation
distribution and replaces this by the selection of appropriate
univariate conditional distributions''. With FCS, an iterative
algorithmic procedure is used to \ldots{} multiple imputation by chained
equations \ldots{} ``The values for imputing the missing data are drawn
from the posterior predictive distribution of the missing data
conditional on the observed data.'' There is no scientific consensus on
the convergence properties of FCS algorithms \citep{taka17}. Some
default techniques in \texttt{mice} might not yield converged states at
all \citep{murr18}. Therefore, algorithmic convergence should be
monitored carefully.

@@

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye
\citep[\(\S\) 6.5.2]{buur18}. Therefore diagnostic evaluation of
convergence would be preferred. However, not known whether MCMC
diagnostics are appropriate for MICE \ldots{} The application of
convergence diagnostics to MI methods has not been systematically
studied \citep{buur18}. How can non-convergence be diagnosed? How many
iterations are sufficient/required? What are the effects of
non-convergence on estimates, predictions and inference? What is an
appropriate non-convergence diagnostic? Are the default number of
iterations sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 1;
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). For each datapoint in \(y_{mis}\) \(M \times T\)
times plausible values are sampled, where \(M\) is the number of
imputations (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, \dots, T\)). The collection of samples between the initial
value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
to as an `imputation chain'.

@@

In iterative algorithmic procedures (Markov chain Monte Carlo methods;
e.g., \texttt{mice} algorithms or Gibbs samplers) non-convergence may be
present as non-stationarity within chains (i.e., trending), or as slow
mixing between chains (i.e., no intermingling). (\textbf{add
example/guidelines visual inspection here?})

The stationarity component of convergence may be evaluated with
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending
within chains.

The mixing component can be assessed with the potential scale reduction
factor \(\widehat{R}\) \citep[a.k.a. `Gelman-Rubin
statistic';][]{gelm92}. With an adapted version of \(\widehat{R}\),
proposed by \citet{veht19}, we might also evaluate the stationarity
component of convergence. This would make \(\widehat{R}\) a general
convergence diagnostic. The application of \(\widehat{R}\) to assess
stationarity has not been thoroughly investigated. Therefore, this study
employs both \(\widehat{R}\) and autocorrelation to investigate
convergence, as recommended by \citep[p.~898]{cowl96}.

@@

\begin{align*}
B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}. 
\end{align*}

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. That is, there is no baseline measure available to
evaluate performance against.The current practice of visually inspecting
imputation chains for signs of non-mixing or non-stationarity can only
diagnose severely pathological cases of non-convergence. We could also
look at distributional characteristics, and plausibility of imputed
values, see \citet{vinknd}. For now, this is outside of the scope of
this study.

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). No hypothesis
was formulated about the performance of \(AC\) as convergence
diagnostic.

@@

\begin{verbatim}
# pseudo-code of simulation 
simulate data 
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    create missingness
    impute the missingness
    compute convergence diagnostics
    perform analysis
    pool results
    compute simulation diagnostics
aggregate convergence and simulation diagnostics
\end{verbatim}

A finite population of \(N=1000\) is simulated to solve a multiple
linear regression problem, where dependent variable \(Y\) is regressed
on independent variables \(X_1\), \(X_2\) and \(X_3\):

@@

\(\widehat{R}\) is computed by implementing Vehtari et al.'s
\citeyearpar{veht19} recommendations. \(AC\) is computed with function
\texttt{stats::acf()}.

@@

We compute bias as the difference between \(\bar{Q}\) and \(Q\).
Confidence interval width (CIW) is defined as the difference between the
lower and upper bound of the 95\% confidence interval (CI95\%) around
\(\bar{Q}\). We compute the CI95\% bounds as

@@

\textbf{Add more info about figure legends and axes.}

\begin{verbatim}
bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{Convergence diagnostics (means)}
\end{figure}

\begin{verbatim}
bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Convergence diagnostics (variances)}
\end{figure}

\begin{verbatim}
bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Convergence diagnostics (regression coefficients)}
\end{figure}

\hypertarget{convergence-diagnostics-1}{%
\subsection{Convergence Diagnostics}\label{convergence-diagnostics-1}}

It is apparent that there is a relation between the number of iterations
per simulation condition (\(T\)) and the convergence diagnostics.
Generally speaking, conditions with longer imputation chains (higher
\(T\)) coincide with less signs of non-convergence
(\(\widehat{R}\)-values approach one, and \(AC\)-values approach zero).
We see more or less equivalent trends for chain means versus chain
variances. We, therefore, only discuss the \(\widehat{R}\) and
\(AC\)-values of chain means.

Figure 2A shows that \(\widehat{R}\)-values generally decrease with
increasing imputation chain lengths. The decline stabilizes somewhere
between the simulation conditions \(T=30\) and \(T=50\). The downward
trend is most pronounced for \(T=3\), and between \(T = 5\) and
\(T = 10\). In the intervening conditions (\(3 \leq T \leq 5\)),
however, we observe a steep increase in \(\widehat{R}\)-values. This
increase implies that non-convergence is under-estimated, and
convergence should not be diagnosed at this point. Based on the
conventional threshold \(\widehat{R} < 1.2\), we would falsely diagnose
convergence at \(T=3\). According to the widely used threshold
\(\widehat{R} < 1.1\), convergence would be diagnosed for conditions
where \(T>9\). If we use the recently recommended threshold
\(\widehat{R} < 1.01\), we would conclude that convergence is reached in
none of the simulation conditions.

The \(AC\)-values displayed in Figure 2B are almost uniformly increasing
as a function of \(T\). The only \(AC\)-value that deviates from this
observation is for \(T=2\). The lowest \(AC\)-value is obtained for
\(T=3\). \%At \(T=5\), the \(AC\)-value reaches the level observed at
\(T=2\). The gradual increase plateaus between \(T=10\) and \(T=30\).
\(AC\)-values in conditions where \(T>70\) are indifferentiable from
zero, indicating stationarity. We see an initial decrease between
\(T=2\) and \(T=3\). Simulation conditions where \(T>5\) have
\(AC\)-values greater than \(T=2\).

According to this diagnostic, none of the simulation conditions show
signs of non-convergence, since we only observe negative or zero
\(AC\)-values. The negative \(AC\)-values, however, indicate increasing
divergence between subsequent samples in conditions where \(T<4\). It
may not be wise to terminating the algorithm at that point, but after
`recovering' from this dip in \(AC\)-values. This would imply \(T=6\) as
the minimal number of iterations.

@@

We use average bias, average confidence interval width, and coverage
rate as performance measures to evaluate \(\widehat{R}\) and \(AC\). We
make the general observation that the simulation diagnostics behave as
theorized for most simulation conditions (bias around zero, stable
confidence interval widths, nominal coverage rate at 95\%).

Figure 3A shows that bias is fairly stable across simulation conditions.
The condition \(T=1\) clearly deviates from this trend with a negative
bias. The bias for \(T=2\) is below average, but within the range of
fluctuations. Conditions where \(T>3\) can be diagnosed as unbiased, but
the average bias across iterations (as shown by a flat Loess line) only
reaches stability at \(T=20\).

CIW is only clearly divergent in the simulation condition \(T=1\), see
Figure 3B. However, under-estimating the variance of \(\bar{Q}\), which
is the case for \(T=3\), may yield spurious inferences. Only conditions
where \(T>3\) are therefore considered sufficient. These conditions have
similar CIWs, but across iterations stability is not established (i.e.,
the Loess line is never completely flat within the 100 simulation
conditions). \%After about twenty-five iterations the average CIW across
all simulation conditions and repetitions is reached 0.927.

The empirical coverage rate across repetitions seems more or less stable
for conditions where \(T>1\), see Figure 3C. We see some over-coverage
at \(T=2\), but that is better than under-estimating the variance of
\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
expected nominal coverage of 95\% \citep{neym34}, namely 95\%. Similar
to the CIWs, there is some trending across iterations (i.e., the Loess
line is never flat).

@@

This note shows that convergence diagnostics \(\widehat{R}\) and \(AC\)
may diagnose convergence of multiple imputation algorithms, but their
performance differs from conventional applications to iterative
algorithmic procedures.\\
\(\widehat{R}\) and autocorrelation indicate that algorithmic
convergence may only be reached after twenty or even forty iterations,
while unbiased, confidence valid estimates estimates may be obtained
with as little as four iterations. These results are in agreement with
the simulation hypothesis: \(\widehat{R}\) over-estimates the severity
of non-convergence when applied to MI procedures. \%This may be due to
the quantity of scientific interest chosen. More `complicated' \(Q\)s
(e.g., higher order effects or variance components) might show bias,
under- or over-coverage at higher \(T\).

According to this simulation study, the recently proposed threshold of
\(\widehat{R}<1.01\) may be too stringent for MI algorithms. Under the
relatively easy missing data problem of the current study, the threshold
was not reached. The other extreme of the \(\widehat{R}\)-thresholds,
the conventionally acceptable \(\widehat{R} <1.2\), may be too lenient
for MI procedures. Applying this threshold to the current data, lead to
falsely diagnosing convergence at \(T = 3\). It appears that the widely
used threshold of \(\widehat{R} < 1.1\) suits MI algorithms the best. We
might, however, also formulate a new threshold, specifically for the
evaluation of MI algorithms. The current study suggests that
\(\widehat{R} < 1.05\) may be implemented, since that is the level at
which the \(\widehat{R}\) stabilize (around \(T = 20\)).

The negative \(AC\)-values obtained in this study show no threat of
non-stationarity. However, initial dip in \(AC\)-values may have
implications for the default number of iterations in \texttt{mice}
(\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
the most appropriate, since this lead to the worst convergence, as
indicated by \(\widehat{R}\) and \(AC\). Under the current
specifications, \(T>20\) would be more appropriate.

The observed dip in AC implies that default maxit value of five
iterations is the worst possible number of iterations. Moreover, the
results of this study imply that assessing the stationarity component of
convergence with \(AC\) might be redundant, since high \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.
\(AC\) would thus not be informative of the convergence of MI
algorithms.

Since this study only considers only an MCAR missingness mechanism,
results may not be extrapolated to other missing data problems. Proper
performance of the convergence diagnostics under MCAR is necessary but
not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
\(AC\) as convergence diagnostics. This is just a proof of concept.

Further research is needed to investigate their performance under clear
violation of convergence, e.g.~dependency between predictors (predictors
with very high correlations). Until then, we have only shown that the
convergence diagnostics can diagnose non-convergence of MI algorithms
that trend towards a converged state. Also for future research,
implement Wernicke diagnostic, and look at developing a convergence
diagnostic for substantive models, and implement a Wald test for
\(AC = 0\).

>>>>

At some point, any scientist conducting statistical analyses will run
into a missing data problem \citep{alli02}. Missingness is problematic
because statistical inference cannot be performed on incomplete data
without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
which may yield wildly invalid results \citep{buur18}. A popular answer
to the ubiquitous problem of missing information is to use the framework
of multiple imputation (MI), proposed by \citet{rubin87}. \textbf{(Call
it mice already, then explain mi and ce sides of mice)} For the
theoretical foundation of MI, see \citet{rubin87}. For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~\citet{buur18}. With MI, missing datapoints are `imputed' (i.e.,
filled in) via an \emph{imputation model} (\textbf{term not required to
understand this paper, so circumvent}) . (\textbf{Move this to mi vs.~ce
part in intro:} The imputation model can be a joint model for all
variables, or several univariate conditional distributions (fully
conditional specification). FCS is more flexible and ``avoids the
problem of specifying an appropriate joint imputation distribution and
replaces this by the selection of appropriate univariate conditional
distributions''.)**

\textbf{(Rewrite: With FCS, an iterative algorithmic procedure is used
to \ldots{} multiple imputation by chained equations \ldots{} ``The
values for imputing the missing data are drawn from the posterior
predictive distribution of the missing data conditional on the observed
data.'')}

There is no scientific consensus on the convergence properties of FCS
algorithms \citep{taka17}. Some default techniques in \texttt{mice}
might not yield converged states at all \citep{murr18}. Therefore,
algorithmic convergence should be monitored carefully.

@@

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye
\citep[\(\S\) 6.5.2]{buur18}. Therefore diagnostic evaluation of
convergence would be preferred. However, not known whether MCMC
diagnostics are appropriate for MICE \ldots{} The application of
convergence diagnostics to MI methods has not been systematically
studied \citep{buur18}. How can non-convergence be diagnosed? How many
iterations are sufficient/required? What are the effects of
non-convergence on estimates, predictions and inference? What is an
appropriate non-convergence diagnostic? Are the default number of
iterations sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 10
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). \textbf{(This is not only notation, but theory too:
For each datapoint in \(y_{mis}\) \(M \times T\) times plausible values
are sampled, where \(M\) is the number of imputations
(\(m = 1, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, \dots, T\)).)} The collection of samples between the initial
value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
to as an `imputation chain'.

@@

\textbf{(Start with current guidelines, then continue with possible
quant measures of the same things from MCMC lit.)} In iterative
algorithmic procedures (Markov chain Monte Carlo methods; e.g.,
\texttt{mice} algorithms or Gibbs samplers) non-convergence may be
present as non-stationarity within chains (i.e., trending), or as slow
mixing between chains (i.e., no intermingling (\textbf{source??})).
(\textbf{add example/guidelines visual inspection here?})

The stationarity component of convergence may be evaluated with
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
\citeyearpar{raft91} procedure to determine (\textbf{the chain length
required to cancel out trending}) the effect of trending within chains.

The mixing component can be assessed with the potential scale reduction
factor \(\widehat{R}\) \citep[a.k.a. `Gelman-Rubin
statistic';][]{gelm92}. With an adapted version of \(\widehat{R}\),
proposed by \citet{veht19}, the stationarity component of convergence
might be evaluated as well. This would make \(\widehat{R}\) a general
convergence diagnostic. The application of \(\widehat{R}\) to assess
stationarity has not been thoroughly investigated. Therefore, this study
employs both \(\widehat{R}\) and autocorrelation to investigate
convergence, as recommended by \citep[p.~898]{cowl96}.

@@

\begin{align*}
B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}. 
\end{align*}

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. That is, there is no baseline measure available to
evaluate performance against. \textbf{(Move above: The current practice
of visually inspecting imputation chains for signs of non-mixing or
non-stationarity can only diagnose severely pathological cases of
non-convergence.)}

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
actual hypothesis.)} No hypothesis was formulated about the performance
of \(AC\) as convergence diagnostic.

@@

A finite population of \(N=1000\) is simulated to solve a multiple
linear regression problem, where dependent variable \(Y\) is regressed
on independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
notation with betas here, and add what the quantity/-ies of scientific
interest is/are!}):

@@

\(\widehat{R}\) (\textbf{of what?? i.e., chain means and chain variances
of each variable}) is computed by implementing Vehtari et al.'s
\citeyearpar{veht19} recommendations. \(AC\) (\textbf{of what? and how
is AC aggregated across imputations? i.e., mean in latest sim}) is
computed with function \texttt{stats::acf()}.

@@

We compute bias as the difference between \(Q\) and \(\bar{Q}\).
Confidence interval width (CIW) is defined as the difference between the
lower and upper bound of the 95\% confidence interval (CI95\%) around
\(\bar{Q}\). We compute the CI95\% bounds as

@@

\textbf{(Add more info about figure legends and axes.)}

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1}

}

\caption{Convergence diagnostics (means)}\label{fig:unnamed-chunk-2}
\end{figure}

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1}

}

\caption{Convergence diagnostics (variances)}\label{fig:unnamed-chunk-3}
\end{figure}

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1}

}

\caption{Convergence diagnostics (regression coefficients)}\label{fig:unnamed-chunk-4}
\end{figure}

\hypertarget{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}{%
\subsection{Convergence Diagnostics (pair convergnce diagnostics with
respective estimates, not split between convergence and simulation
diagnostics)}\label{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}}

\textbf{(Add different percentages missingness??)}

It is apparent that there is a relation between the number of iterations
per simulation condition (\(T\)) and the convergence diagnostics.
Generally speaking, conditions with longer imputation chains (higher
\(T\)) coincide with less signs of non-convergence
(\(\widehat{R}\)-values approach one, and \(AC\)-values approach zero).

Figure 2A shows that \(\widehat{R}\)-values generally decrease with
increasing imputation chain lengths. The decline stabilizes somewhere
between the simulation conditions \(T=30\) and \(T=50\). The downward
trend is most pronounced for \(T=3\), and between \(T = 5\) and
\(T = 10\). In the intervening conditions (\(3 \leq T \leq 5\)),
however, we observe a steep increase in \(\widehat{R}\)-values. This
increase implies that non-convergence is under-estimated, and
convergence should not be diagnosed at this point (\textbf{is that
true???}). Based on the conventional threshold \(\widehat{R} < 1.2\), we
would falsely diagnose convergence at \(T=3\). According to the widely
used threshold \(\widehat{R} < 1.1\), convergence would be diagnosed for
conditions where \(T>9\). If we use the recently recommended threshold
\(\widehat{R} < 1.01\), we would conclude that convergence is reached in
none of the simulation conditions.

The \(AC\)-values displayed in Figure 2B are almost uniformly increasing
as a function of \(T\). The only \(AC\)-value that deviates from this
observation is for \(T=2\). The lowest \(AC\)-value is obtained for
\(T=3\). \%At \(T=5\), the \(AC\)-value reaches the level observed at
\(T=2\). The gradual increase plateaus between \(T=10\) and \(T=30\).
\(AC\)-values in conditions where \(T>70\) are indifferentiable from
zero, indicating stationarity. We see an initial decrease between
\(T=2\) and \(T=3\). Simulation conditions where \(T>5\) have
\(AC\)-values greater than \(T=2\) (\textbf{weird wording}).

According to this diagnostic, none of the simulation conditions show
signs of non-convergence, since we only observe negative or zero
\(AC\)-values. The negative \(AC\)-values, however, indicate increasing
divergence (\textbf{move above!}) between subsequent samples in
conditions where \(T<4\). It may not be wise (\textbf{why? and how would
we know? e.g., wider distrubutions? didn't seem like it\ldots{}}) to
terminating the algorithm at that point, but after `recovering' from
this dip in \(AC\)-values. This would imply \(T=6\) as the minimal
number of iterations.

@@

We use average bias, average confidence interval width, and coverage
rate as performance measures to evaluate \(\widehat{R}\) and \(AC\)
(\textbf{not the goal! we want to show the effects of non-conv, not just
see how to diagnose it.}). We make the general observation that the
simulation diagnostics behave as theorized for most simulation
conditions (bias around zero, stable confidence interval widths, nominal
coverage rate at 95\%).

Figure 3A shows that bias is fairly stable across simulation conditions.
The condition \(T=1\) clearly deviates from this trend with a negative
bias. The bias for \(T=2\) is below average, but within the range of
fluctuations. Conditions where \(T>3\) can be diagnosed as unbiased, but
the average bias across iterations (as shown by a flat Loess line) only
reaches stability at \(T=20\) (\textbf{Loess might not be relevant,
check and remove accordingly!}).

CIW is only clearly divergent in the simulation condition \(T=1\), see
Figure 3B. However, under-estimating the variance of \(\bar{Q}\), which
is the case for \(T=3\), may yield spurious inferences (\textbf{move
above}). Only conditions where \(T>3\) are therefore considered
sufficient. (\textbf{Possibly irrelevant Loess: These conditions have
similar CIWs, but across iterations stability is not established (i.e.,
the Loess line is never completely flat within the 100 simulation
conditions).})

The empirical coverage rate across repetitions seems more or less stable
for conditions where \(T>1\), see Figure 3C. We see some over-coverage
at \(T=2\), but that is better than under-estimating the variance of
\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
expected nominal coverage of 95\% \citep{neym34} (\textbf{not
anymore??}). (\textbf{Loess: Similar to the CIWs, there is some trending
across iterations (i.e., the Loess line is never flat)}).

@@

This note shows that convergence diagnostics \(\widehat{R}\) and \(AC\)
may diagnose convergence of multiple imputation algorithms, but their
performance differs from conventional applications to iterative
algorithmic procedures. (\textbf{nope! it shows that MICE can lead to
correct outcomes when they have not converged accroding to two common
conv diags. This may be due to the measures (e.g., assumption of
overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more
complex RQs). Add what \%miss has to do with it.})\\
\(\widehat{R}\) and autocorrelation indicate that algorithmic
convergence may only be reached after twenty or even forty iterations,
while unbiased, confidence valid estimates estimates may be obtained
with as little as four iterations. These results are in agreement with
the simulation hypothesis: \(\widehat{R}\) over-estimates the severity
of non-convergence when applied to MI procedures. \%This may be due to
the quantity of scientific interest chosen. More `complicated' \(Q\)s
(e.g., higher order effects or variance components) might show bias,
under- or over-coverage at higher \(T\).

According to this simulation study, the recently proposed threshold of
\(\widehat{R}<1.01\) may be too stringent for MI algorithms.
(\textbf{This is only one of the goals: to give applied researchers a
diagnostic to indicate that they should keep iterating. The other is the
default in mice and other software packages, and yet another is \ldots{}
i forgot}) Under the relatively easy missing data problem of the current
study, the threshold was not reached. The other extreme of the
\(\widehat{R}\)-thresholds, the conventionally acceptable
\(\widehat{R} <1.2\), may be too lenient for MI procedures. Applying
this threshold to the current data, lead to falsely diagnosing
convergence at \(T = 3\) (\textbf{because it goes up after, not because
it is not converged enough}). It appears that the widely used threshold
of \(\widehat{R} < 1.1\) suits MI algorithms the best. We might,
however, also formulate a new threshold, specifically for the evaluation
of MI algorithms. The current study suggests that \(\widehat{R} < 1.05\)
may be implemented, since that is the level at which the \(\widehat{R}\)
stabilize (around \(T = 20\)) (\textbf{Not necessary for this Q, but
maybe for more complicated Qs}).

The negative \(AC\)-values obtained in this study show no threat of
non-stationarity. However, initial dip in \(AC\)-values may have
implications for the default number of iterations in \texttt{mice}
(\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
the most appropriate, since this lead to the worst convergence
(\textbf{nope, only for Rh, not AC}), as indicated by \(\widehat{R}\)
and \(AC\). Under the current specifications, \(T>20\) would be more
appropriate.

The observed dip in AC implies that default maxit value of five
iterations is the worst possible number of iterations. Moreover, the
results of this study imply that assessing the stationarity component of
convergence with \(AC\) might be redundant, since high \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains
(\textbf{add above!!}). \(AC\) would thus not be informative of the
convergence of MI algorithms.

(\textbf{Add above, not here because it was a known limitation: Since
this study only considers only an MCAR missingness mechanism, results
may not be extrapolated to other missing data problems. Proper
performance of the convergence diagnostics under MCAR is necessary but
not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
\(AC\) as convergence diagnostics. This is just a proof of concept.})

Further research is needed to investigate their performance under clear
violation of convergence, e.g.~dependency between predictors (predictors
with very high correlations). Until then, we have only shown that the
convergence diagnostics can diagnose non-convergence of MI algorithms
that trend towards a converged state. Also for future research, look at
developing a convergence diagnostic for substantive models, and
implement a Wald test for \(AC = 0\).
